---
title: "Validation script"
author: "Johan Lundin Kleberg"
date: "2025-05-13"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## R Markdown

Load necessary libraries
```{r}

library("FactoMineR")
library(dplyr)
library(openxlsx)
library("factoextra")
library(vegan)
library(tidyverse)
library(cowplot)
library(colorspace)
library(ggrepel)
library(lme4)
library(DHARMa)
library(fitdistrplus)
library("ggpubr")
library(emmeans)
library(TSdist)
library(gtools)
library(stringr)
```

## Setup a function  

This function will 1) run the MCA, and 2) return the data used to create plot 4 in the published manuscript. The code is adapted so that the numbers of factors in the MCA can be specified by the user. 

```{r}
xdata1.factor.nr<- function(xdata, n.factors) {
  
  res.mca<-MCA(xdata,quanti.sup=c(5,6), quali.sup=c(1:4), ncp = n.factors, graph = TRUE) #Run the MCA. Set 'graph' to TRUE to plot the output
  #res.mca$eig # Extract the eigenvalues of the first axes
  
  
  results <- catdes(xdata, num.var=1) # Extract the FoCs associated more than expected by chance with each utterance type. 
  # Plot the MCA results (Figure 1)
  name <- list(name = levels(xdata$Combination)) # Keep the name of the utterances we want to display on the plot
  #################################################### Extract distances between utterances ####################################################
  
  coord <- as.data.frame(res.mca$quali.sup$coord[1:26,]) # Extract the 5-dimension coordinates of each utterance type
  distances <- as.matrix(vegdist(coord, method="euclidian")) # Measure the distance between each utterance type
  
  # Create a function to assess whether a single call is part of a combination
  is_in_combination <- function(cri, combination) {
    cri_parts <- unlist(strsplit(cri, "-"))
    any(cri_parts %in% unlist(strsplit(combination, "-")))
  }
  
  # Extract, for each possible pair of utterance types of the dataset, the distance between the two utterances and their relationship (the utterances are two single calls, two combinations, or one signal call is a constituent or not of the other combination) 
  newdata <- data.frame(matrix(nrow = 351))
  processed_pairs <- list()
  k=1
  for (i in 1:nrow(distances)){
    for (j in 1:ncol(distances)){
      
      pair <- sort(c(rownames(distances)[i], colnames(distances)[j]))
      pair_string <- paste(pair, collapse = "_")
      if (!(pair_string %in% processed_pairs)) {
        
        if ((nchar(rownames(distances)[i])==nchar(colnames(distances)[j])) & (nchar(rownames(distances)[i])==2)) {
          newdata$dist[k] <- distances[i,j]
          newdata$call_1[k] <- colnames(distances)[j]
          newdata$call_2[k] <- rownames(distances)[i]
          newdata$pair[k] <- pair_string
          newdata$comb[k] <- "single"
          k <- k+1
        } else if ((nchar(rownames(distances)[i])==nchar(colnames(distances)[j])) & (nchar(rownames(distances)[i])==5)) {
          newdata$dist[k] <- distances[i,j]
          newdata$call_1[k] <- colnames(distances)[j]
          newdata$call_2[k] <- rownames(distances)[i]
          newdata$pair[k] <- pair_string
          newdata$comb[k] <- "two_comb"
          k <- k+1
        } else if (is_in_combination(rownames(distances)[i], colnames(distances)[j])==TRUE) {
          newdata$dist[k] <- distances[i,j]
          newdata$call_1[k] <- colnames(distances)[j]
          newdata$call_2[k] <- rownames(distances)[i]
          newdata$pair[k] <- pair_string
          newdata$comb[k] <- "within_comb"
          k <- k+1
        } else {
          newdata$dist[k] <- distances[i,j]
          newdata$call_1[k] <- colnames(distances)[j]
          newdata$call_2[k] <- rownames(distances)[i]
          newdata$pair[k] <- pair_string
          newdata$comb[k] <- "outside_comb"
          k <- k+1
        }
        processed_pairs <- c(processed_pairs, pair_string) # To keep track of which pairs have been processed already
      }
    }
  }
  
  ############################## The meaning of AB is derived from the meaning of A and B  ###################
  
  
  xdata_1 <- rbind(newdata[which(newdata$comb=="within_comb"),], newdata[which(newdata$comb=="outside_comb"),]) # Substract the dataset and keep all the pairs one combination - one single type
  xdata_1$comb <- as.factor(xdata_1$comb)
  xdata_1$pair <- as.factor(xdata_1$pair)
  
  # Indicate, for each pair, the name of the combination type
  xdata_1$comb2 <- apply(xdata_1, 1, function(row) {
    if(nchar(row["call_1"]) > nchar(row["call_2"])) {
      return(row["call_1"])
    } else {return(row["call_2"])}
  })
  xdata_1$comb2 <- as.factor(xdata_1$comb2)
  
  out <- list()
  out[["xdata_1"]] <- xdata_1
  out[["results"]] <- results #FCAs associated with each utterance type
  
  return(out)
  
}
```

## Load the data

```{r}

#Load data
xdata <- read.delim("xdata.txt", header=TRUE, stringsAsFactors = TRUE)
xdata <- xdata[,-c(1)] # For clearer results, we remove the recording names from the analysis
```


## Run the analysis with a MCA with 4 factors. This results in 2 instead of 4 significant contrasts
```{r}

#Number of factors
n.factors <- 4
dat <- xdata1.factor.nr(xdata, n.factors = n.factors)

# Create the full model and check model assumptions
model1 <- lm(dist~comb*comb2, data=dat[["xdata_1"]])

model1_emm <- emmeans(model1, pairwise~comb|comb2) # Post hoc analysis: which combination is closer to its constituting calls than to other single calls of the repertoire

# Plot the results (Figure 4)
d <- as.data.frame(model1_emm$emmeans)
e <- as.data.frame(model1_emm$contrasts)

#ADD A COLUMN WITH ADJUSTED P-VALUES FOR COMPARISION. THIS REDUCES THE NUMBER OF SIGNIFICANT CONTRASTS 
e$p.value.adj <- p.adjust(e$p.value, method = "fdr")

e_4factors <- e #Save for comparisons with other analyses

# Plot unadjusted p-values as in the manuscript. Change to
# p.value.variable <- "p.adjust" #to plot adjusted p-values instead
p.value.variable <- "p.value"


significant_contrasts <- e[e[[p.value.variable]] < 0.05,]

#Create a label with stars corresponding to p-value to use in plot 4
star.labels <- array(dim = nrow(significant_contrasts))
star.labels[significant_contrasts[[p.value.variable]]< 0.05] <- "*"
star.labels[significant_contrasts[[p.value.variable]]< 0.01] <- "**"
star.labels[significant_contrasts[[p.value.variable]]< 0.001] <- "***"



#Adjusted code for plot 4 which adapts to differences in number of significant contrasts
fig4 <- ggplot(d, aes(x = comb2, y = emmean)) +
  geom_point(aes(color = comb), position = position_dodge(width = 0.45), size = 3) +
  geom_errorbar(aes(ymin = lower.CL, ymax = upper.CL, color = comb), position = position_dodge(width = 0.45), width = 0.25) +
  labs(x = "", y = "Euclidean distance", fill = "") +
  scale_color_manual(name=NULL, values = c("#CC79A7","#56B4E9"), label=c("Outside", "Within")) +
  scale_x_discrete(name="", labels=c("Grunt_Yelp", "High-hoot_Low-hoot", "High-hoot_Peep-yelp", "Low-hoot_High-hoot", "Peep_Grunt", "Peep_Peep-yelp", "Peep_Whistle", "Peep_Yelp", "Peep-yelp_Grunt", "Peep-yelp_High-hoot", "Peep-yelp_Peep", "Peep-yelp_Whistle", "Peep-yelp_Yelp", "Whistle_High-hoot", "Whistle_Peep", "Whistle_Peep-yelp", "Yelp_Grunt", "Yelp_Peep", "Yelp_Peep-yelp"))+
  geom_text(aes(x = comb2, y = 1.7), data=significant_contrasts, label = star.labels, size=7)+
  theme_minimal_hgrid(12, rel_small = 1)+
  theme(
    legend.position = "right",
    legend.justification = "right",
    legend.text = element_text(size = 9),
    legend.box.spacing = unit(0, "pt"), axis.text.x = element_text(angle = 45, hjust = 1, size=9)
  )
title <- paste0("Significant Contrasts (", n.factors, " MCA factors)")
fig4 <- fig4 + ggtitle(title)+ theme(plot.title = element_text(size = 16, face = "bold", hjust = 0.5))

print(fig4)

```

## Plot estimated marginal means (including undadjusted and adjusted p-values) for an analysis with 4 factors
```{r}

print(e)

```



## Run the analysis with a MCA with 6 factors
```{r}

#Number of factors
n.factors <- 6
dat <- xdata1.factor.nr(xdata, n.factors = n.factors)

# Create the full model and check model assumptions
model1 <- lm(dist~comb*comb2, data=dat[["xdata_1"]])

model1_emm <- emmeans(model1, pairwise~comb|comb2) # Post hoc analysis: which combination is closer to its constituting calls than to other single calls of the repertoire

# Plot the results (Figure 4)
d <- as.data.frame(model1_emm$emmeans)
e <- as.data.frame(model1_emm$contrasts)

#ADD A COLUMN WITH ADJUSTED P-VALUES FOR COMPARISION. THIS REDUCES THE NUMBER OF SIGNIFICANT CONTRASTS 
e$p.value.adj <- p.adjust(e$p.value, method = "fdr")

e_6factors <- e #Save for comparisons with other analyses


# Plot unadjusted p-values as in the manuscript. Change to
# p.value.variable <- "p.adjust" #to plot adjusted p-values instead
p.value.variable <- "p.value"


significant_contrasts <- e[e[[p.value.variable]] < 0.05,]

#Create a label with stars corresponding to p-value to use in plot 4
star.labels <- array(dim = nrow(significant_contrasts))
star.labels[significant_contrasts[[p.value.variable]]< 0.05] <- "*"
star.labels[significant_contrasts[[p.value.variable]]< 0.01] <- "**"
star.labels[significant_contrasts[[p.value.variable]]< 0.001] <- "***"



#Adjusted code for plot 4 which adapts to differences in number of significant contrasts
fig4 <- ggplot(d, aes(x = comb2, y = emmean)) +
  geom_point(aes(color = comb), position = position_dodge(width = 0.45), size = 3) +
  geom_errorbar(aes(ymin = lower.CL, ymax = upper.CL, color = comb), position = position_dodge(width = 0.45), width = 0.25) +
  labs(x = "", y = "Euclidean distance", fill = "") +
  scale_color_manual(name=NULL, values = c("#CC79A7","#56B4E9"), label=c("Outside", "Within")) +
  scale_x_discrete(name="", labels=c("Grunt_Yelp", "High-hoot_Low-hoot", "High-hoot_Peep-yelp", "Low-hoot_High-hoot", "Peep_Grunt", "Peep_Peep-yelp", "Peep_Whistle", "Peep_Yelp", "Peep-yelp_Grunt", "Peep-yelp_High-hoot", "Peep-yelp_Peep", "Peep-yelp_Whistle", "Peep-yelp_Yelp", "Whistle_High-hoot", "Whistle_Peep", "Whistle_Peep-yelp", "Yelp_Grunt", "Yelp_Peep", "Yelp_Peep-yelp"))+
  geom_text(aes(x = comb2, y = 1.7), data=significant_contrasts, label = star.labels, size=7)+
  theme_minimal_hgrid(12, rel_small = 1)+
  theme(
    legend.position = "right",
    legend.justification = "right",
    legend.text = element_text(size = 9),
    legend.box.spacing = unit(0, "pt"), axis.text.x = element_text(angle = 45, hjust = 1, size=9)
  )
title <- paste0("Significant Contrasts (", n.factors, " MCA factors)")
fig4 <- fig4 + ggtitle(title)+ theme(plot.title = element_text(size = 16, face = "bold", hjust = 0.5))

print(fig4)

```



## Run the analysis with a MCA with 5 factors (as in the manuscript)
```{r}

#Number of factors
n.factors <- 5
dat <- xdata1.factor.nr(xdata, n.factors = n.factors)

# Create the full model and check model assumptions
model1 <- lm(dist~comb*comb2, data=dat[["xdata_1"]])

model1_emm <- emmeans(model1, pairwise~comb|comb2) # Post hoc analysis: which combination is closer to its constituting calls than to other single calls of the repertoire

# Plot the results (Figure 4)
d <- as.data.frame(model1_emm$emmeans)
e <- as.data.frame(model1_emm$contrasts)

#ADD A COLUMN WITH ADJUSTED P-VALUES FOR COMPARISION. THIS REDUCES THE NUMBER OF SIGNIFICANT CONTRASTS 
e$p.value.adj <- p.adjust(e$p.value, method = "fdr")

e_5factors <- e #Save for comparisons with other analyses


# Plot unadjusted p-values as in the manuscript. Change to
# p.value.variable <- "p.adjust" #to plot adjusted p-values instead
p.value.variable <- "p.value"


significant_contrasts <- e[e[[p.value.variable]] < 0.05,]

#Create a label with stars corresponding to p-value to use in plot 4
star.labels <- array(dim = nrow(significant_contrasts))
star.labels[significant_contrasts[[p.value.variable]]< 0.05] <- "*"
star.labels[significant_contrasts[[p.value.variable]]< 0.01] <- "**"
star.labels[significant_contrasts[[p.value.variable]]< 0.001] <- "***"



#Adjusted code for plot 4 which adapts to differences in number of significant contrasts
fig4 <- ggplot(d, aes(x = comb2, y = emmean)) +
  geom_point(aes(color = comb), position = position_dodge(width = 0.45), size = 3) +
  geom_errorbar(aes(ymin = lower.CL, ymax = upper.CL, color = comb), position = position_dodge(width = 0.45), width = 0.25) +
  labs(x = "", y = "Euclidean distance", fill = "") +
  scale_color_manual(name=NULL, values = c("#CC79A7","#56B4E9"), label=c("Outside", "Within")) +
  scale_x_discrete(name="", labels=c("Grunt_Yelp", "High-hoot_Low-hoot", "High-hoot_Peep-yelp", "Low-hoot_High-hoot", "Peep_Grunt", "Peep_Peep-yelp", "Peep_Whistle", "Peep_Yelp", "Peep-yelp_Grunt", "Peep-yelp_High-hoot", "Peep-yelp_Peep", "Peep-yelp_Whistle", "Peep-yelp_Yelp", "Whistle_High-hoot", "Whistle_Peep", "Whistle_Peep-yelp", "Yelp_Grunt", "Yelp_Peep", "Yelp_Peep-yelp"))+
  geom_text(aes(x = comb2, y = 1.7), data=significant_contrasts, label = star.labels, size=7)+
  theme_minimal_hgrid(12, rel_small = 1)+
  theme(
    legend.position = "right",
    legend.justification = "right",
    legend.text = element_text(size = 9),
    legend.box.spacing = unit(0, "pt"), axis.text.x = element_text(angle = 45, hjust = 1, size=9)
  )
title <- paste0("Significant Contrasts (", n.factors, " MCA factors)")
fig4 <- fig4 + ggtitle(title)+ theme(plot.title = element_text(size = 16, face = "bold", hjust = 0.5))

print(fig4)

```



## Run statistical analyses with data from the five factor MCA

# Step 1: Run the full interaction model (main and interaction effects) 
```{r}
model1 <- lm(dist~comb*comb2, data=dat[["xdata_1"]])


```


# Step 1: Compare this to an intercept only model. 
```{r}

model_null <- lm(dist~1, data=dat[["xdata_1"]]) # Create a null model and compare it to the full model using a likelihood ratio test 

anova(model1, model_null) # The difference if significative


```

# Step 2: Compare the full model to alternative models. 

First, as in the paper, compare to model_comb which only includes a main effect of "comb" meaning that 
distance depends on whether the two calls were made together or on separate occasions. Note that this comparison does not compare the full model to the next most complex model since mod_comb does not include a main effect of call. The analysis yields a significant effect but does not test the interaction shown in Figure 4

```{r}
# Create two reduced models and compare them to the full model using a likelihood ratio test
model_comb <- lm(dist~comb, data=dat[["xdata_1"]]) 
anova(model_comb, model1) # The difference is significative

model_comb2 <- lm(dist~comb2, data=dat[["xdata_1"]])
anova(model_comb2, model1) # The difference is significative - this is the model reported 

```

Second, as in the paper, compare to model_comb which only includes a main effect of "comb2" meaning that 
distance depends on the type of call. Note that this comparison does not compare the full model to the next most complex model since mod_comb does not include a main effect of "comb" (i.e, whether the distance is calculated within or between pairs). The analysis yields a significant effect but does not test the interaction shown in Figure 4

```{r}
# Create two reduced models and compare them to the full model using a likelihood ratio test

model_comb2 <- lm(dist~comb2, data=dat[["xdata_1"]])
anova(model_comb2, model1) # The difference is significative - this is the model reported 

```


Now, compare the full model to the next most complex model with a likelihood ratio test. This is the standard way to test whether the interaction effect is significant.
The interaction effect is NOT significant,

```{r}
# Create two reduced models and compare them to the full model using a likelihood ratio test

model_without_interactions <- lm(dist~comb2+comb, data=dat[["xdata_1"]])
anova(model_without_interactions, model1) # The difference is significative - this is the model reported 

```


## Plot the occurrences of call combinations identified as compositional (present in the df "significant_contrasts).
Examples of call combinations claimed to be compositional are very rare in the data set. Very few individuals produce them on more than one occasion.

Note that only "pe-wi" "ye-gr" give significant results a four factor MCA. "ye-gr" is the only call with a significant distance difference from its constituents outside the combination with a liberal correction for multiple comparisons in the 5 factor analysis.

```{r, fig.width=15, fig.height=5}
# Read the data again to access the 'Recording' variable
xdata2 <- read.delim("xdata.txt", header=TRUE, stringsAsFactors = TRUE)
xdata2$day <- substr(x = xdata2$Recording, start = 1, stop = 8)
xdata2$group_day <- paste0(xdata2$Group, xdata$day)
targets <- significant_contrasts$comb2


total.occurences <- xdata2 %>% group_by(Combination) %>% summarize(n = n())
total.by.id <- xdata2 %>% group_by(Caller) %>% summarize(n = n())


target.data <- filter(xdata2, Combination %in% targets)

title <- paste0("Number of call combinations")
observed.calls.plot <- ggplot(target.data, aes(x = Caller, fill = day)) +geom_bar() + 
  facet_wrap(vars(Combination), ncol = 1)+ scale_y_continuous(limits = c(0,4))+
  ggtitle(title) + theme(plot.title = element_text(size = 16, face = "bold", hjust = 0.5))

print(observed.calls.plot)
```
## Extract factor loadings by FOC

# Run the MCA with 5 factors. Extract factor loadings
```{r}
#Run the MCA with 5 factors. Incpet the factor loadings
res.mca<-MCA(xdata,quanti.sup=c(5,6), quali.sup=c(1:4), ncp = 5, graph = FALSE) #Run the MCA. Set 'graph' to TRUE to plot the output

contrib <- res.mca[["var"]][["contrib"]]
contrib <- round(contrib,2)
factor.belonging <- max.col(contrib)

foc <- row.names(contrib)
foc.contrib <- data.frame(contrib)
foc.contrib$factor.belonging <- factor.belonging
foc.contrib$foc <- foc
row.names(foc.contrib) <- NULL


```


#Incpect the FOCs associated with factor 5

```{r}
foc.contrib <- foc.contrib[order(foc.contrib$Dim.5, decreasing = TRUE),]
foc.contrib$foc[foc.contrib$factor.belonging == 5]

```



